One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-02-06 paper/paper.tex
paper accepted, minor corrections

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
 an imposed radial temperature gradient. This might apply, for
 example, to the outer parts of protoplanetary discs. We use linear
 density wave theory 
 to show that non-axisymmetric perturbations generally do not conserve
 their angular momentum in the presence of a forced temperature
 gradient. This implies an exchange of angular momentum 
 between linear perturbations and the background disc. In particular, 
 when the disturbance is a low-frequency 
 trailing wave and the disc temperature decreases outwards, this 
 interaction is unstable and leads to the growth of the wave.  
 We demonstrate this phenomenon through numerical hydrodynamic
 simulations of locally isothermal discs in 2D using the FARGO
 code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
 radially structured discs with a self-gravitating region which remains
 stable in the absence of a temperature gradient. However,
 when a temperature gradient is imposed we observe exponential growth of a one-armed  
 spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
 outside the bulk of the spiral arm, resulting in a nearly-stationary
 one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by a forced temperature gradient does 
 not explicitly require disc self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}

@@

% planet is popular answer for spiral structure
The attractive explanation for asymmetries in circumstellar discs is
disc-planet interaction. In particular, spiral structures  
naturally arise from the gravitational interaction between a
planet and the gaseous protoplanetary disc it is embedded in
\citep[see, e.g.][for a recent review]{baruteau13b}. Thus, the
presence of spiral arms in circumstellar discs could be signposts of 
planet formation \citep{juhasz14}.

% but also natural in GI.
Spiral arms are also {\bf characteristic of} global gravitational 
instability (GI) {\bf in} differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale 
spiral arms can provide significant angular momentum transport
necessary for mass accretion \citep{lynden-bell72,
  papaloizou91,balbus99,lodato04}, and spiral structures due to GI are
potentially observable with the Atacama Large
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}. GI can
be expected in the earliest stages of circumstellar disc formation 
\citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
outer parts of the disc \citep{rafikov05,matzner05,kimura12}.

Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in 
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
In the `SLING' mechanism proposed by \cite{shu90}, a $m=1$ {\bf
  gravitational} instability arises from the motion of the
central star induced by the one-armed perturbation, and requires a
massive disc \citep[the former may have observable consequences,
][]{michael10}.

{\bf
  In this work we identify a new mechanism that leads to the growth of
  one-armed spirals in astrophysical discs. We show
  that when the disc temperature is prescribed (called locally isothermal
  discs), the usual statement of the conservation of angular momentum
  for linear perturbations acquires a source term proportional to the
  temperature gradient. This permits angular momentum exchange between
  linear perturbations and the background disc. This `background
  torque' can destabilise low-frequency non-axisymmetric
  trailing waves when the disc temperature decreases outwards.       
}

{\bf We employ direct hydrodynamic simulations using three different
  grid-based codes to demonstrate how this `background torque' 
  can lead to the growth of one-armed spirals in radially structured, 
  self-gravitating discs. This is despite the fact that our disc
  models do not meet the requirements for the `SLING' 
  mechanism. Although our numerical simulations consider
  self-gravitating discs, this `background torque' is generic for
  locally isothermal discs and its existence does not require
  disc self-gravity. Thus, the destabilisation effect we
  describe should also be applicable to non-self-gravitating discs.  
}

This paper is organised as follows. In \S\ref {model} we describe the system of interest
and list the governing equations. {\bf In \S\ref{wkb} we use linear theory  to show
  how a fixed temperature profile 
  can destabilise non-axisymmetric waves in discs.} \S\ref{methods} describes
the {\bf numerical setup and} hydrodynamic codes {\bf we use to 
  demonstrate the growth of one-armed spirals due to an imposed
  temperature gradient.} Our {\bf
  simulation} results are presented in  \S\ref{results2d} for two-dimensional (2D) {\bf
  discs} and in \S\ref{results3d} for three-dimensional (3D) {\bf discs}, and we further
discuss them in \S\ref{discussions}. We summarise in \S\ref{summary} 
with some speculations for future work.

@@

{\bf
\section*{Acknowledgments}
I thank K. Kratter, Y. Wu and A. Youdin for valuable discussions, and the
anonymous referee for comments that significantly improved this paper. All
computations were performed on the El Gato cluster at the University
of Arizona. This material is based upon work supported by the National
Science Foundation under Grant No. 1228509. 
}
\bibliographystyle{mn2e}
\bibliography{ref}

>>>>

\maketitle
\begin{abstract} 
   We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
 an imposed radial temperature gradient. This might apply, for
 example, to the outer parts of protoplanetary discs. We use linear
 density wave theory 
 to show that non-axisymmetric perturbations generally do not conserve
 their angular momentum in the presence of a forced temperature
 gradient. This implies an exchange of angular momentum 
 between linear perturbations and the background disc. In particular, 
 when the disturbance is a low-frequency 
 trailing wave and the disc temperature decreases outwards, this 
 interaction is unstable and leads to the growth of the wave.  
 We demonstrate this phenomenon through numerical hydrodynamic
 simulations of locally isothermal discs in 2D using the FARGO
 code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
 radially structured discs with a self-gravitating region which remains
 stable in the absence of a temperature gradient. However,
 when a temperature gradient is imposed we observe exponential growth of a one-armed  
 spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
 outside the bulk of the spiral arm, resulting in a nearly-stationary
 one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by a forced temperature gradient does 
 not explicitly require disc self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs.

\end{abstract}

@@

% planet is popular answer for spiral structure
The attractive explanation for asymmetries in circumstellar discs is
disc-planet interaction. In particular, spiral structures  
naturally arise from the gravitational interaction between a
planet and the gaseous protoplanetary disc it is embedded in
\citep[see, e.g.][for a recent review]{baruteau13b}. Thus, the
presence of spiral arms in circumstellar discs could be signposts of 
planet formation \citep[but see][]{juhasz14}.

% but also natural in GI.
Spiral arms are also  characteristic of global gravitational 
instability (GI)  in differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale 
spiral arms can provide significant angular momentum transport
necessary for mass accretion \citep{lynden-bell72,
  papaloizou91,balbus99,lodato04}, and spiral structures due to GI are
potentially observable with the Atacama Large
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}. GI can
be expected in the earliest stages of circumstellar disc formation 
\citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
outer parts of the disc \citep{rafikov05,matzner05,kimura12}.

Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in 
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
In the `SLING' mechanism proposed by \cite{shu90}, an $m=1$ 
  gravitational instability arises from the motion of the
central star induced by the one-armed perturbation, and requires a
massive disc \citep[the former may have observable consequences,
][]{michael10}.

  In this work we identify a new mechanism that leads to the growth of
  one-armed spirals in astrophysical discs. We show
  that when the disc temperature is prescribed (called locally isothermal
  discs), the usual statement of the conservation of angular momentum
  for linear perturbations acquires a source term proportional to the
  temperature gradient. This permits angular momentum exchange between
  linear perturbations and the background disc. This `background
  torque' can destabilise low-frequency non-axisymmetric
  trailing waves when the disc temperature decreases outwards.

 We employ direct hydrodynamic simulations using three different
  grid-based codes to demonstrate how this `background torque' 
  can lead to the growth of one-armed spirals in radially structured, 
  self-gravitating discs. This is despite the fact that our disc
  models do not meet the requirements for the `SLING' 
  mechanism. Although our numerical simulations consider
  self-gravitating discs, this `background torque' is generic for
  locally isothermal discs and its existence does not require
  disc self-gravity. Thus, the destabilisation effect we
  describe should also be applicable to non-self-gravitating discs.

This paper is organised as follows. In \S\ref {model} we describe the system of interest
and list the governing equations. In \S\ref{wkb} we use linear theory  to show
  how a fixed temperature profile 
  can destabilise non-axisymmetric waves in discs. \S\ref{methods} describes
the numerical setup and hydrodynamic codes we use to 
  demonstrate the growth of one-armed spirals due to an imposed
  temperature gradient. Our
  simulation results are presented in  \S\ref{results2d} for two-dimensional (2D) 
  discs and in \S\ref{results3d} for three-dimensional (3D) discs, and we further
discuss them in \S\ref{discussions}. We summarise in \S\ref{summary}  
with some speculations for future work.

@@

\section*{Acknowledgments}
I thank K. Kratter, Y. Wu and A. Youdin for valuable discussions, and the
anonymous referee for comments that significantly improved this paper. All
computations were performed on the El Gato cluster at the University
of Arizona. This material is based upon work supported by the National
Science Foundation under Grant No. 1228509.

\bibliographystyle{mn2e}
\bibliography{ref}

### 2015-02-06 paper/results.tex
paper accepted, minor corrections

\subsection{Reference run}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc {\bf is } 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibly due to non-linear interaction of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

Fig. \ref{2d_angmom} shows the evolution of disc angular momentum
components. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum, {\bf which indicates it is a
  low-frequency mode.}   
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-armed spiral to the background disc.

@@

\subsection{Dependence on the imposed temperature profile}
{\bf We show that the growth of the $m=1$ spiral is
  associated with the imposed temperature gradient  by performing a 
  series of simulations with $q\in[0,1]$.  However, to maintain similar Toomre $Q$
  profiles, we adjust the surface density power-law index such
  that $s = (3+q)/2$.  For clarity these simulations are initialised
  with $m=1$ perturbations only.  
}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq}   
  \caption{Evolution of the $m=1$ spiral amplitude as a function of
    the {\bf negative of the imposed temperature gradient} $q$. The maximum value of the
    $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
    \label{fargo_varq}} 
\end{figure}

We plot growth rates of the $m=1$ mode as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a 
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates 
{\bf As the background torque is proportional to $q$
  (Eq. \ref{theoretical_rate})}, 
this {\bf indicates} that the temperature gradient is responsible for the
development of the one-armed spirals observed in our simulations.

@@

\subsection{Properties of the $m=1$ spiral and its growth}\label{fargo_m1}
{\bf Here we analyse the $q=1$ case in Fig. \ref{fargo_varq_growth} in
  more detail.  
} 
Fig. \ref{2d_fargo_viz} shows a snapshot of the $m=1$ surface
density {\bf of this run}. 
By measuring the $m=1$ surface density amplitude and its
pattern speed, we obtain a co-rotation radius and growth rate 
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
This one-armed spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We calculate $k$ numerically and plot
its normalised value in Fig. \ref{fargo_wavenumber}. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply {\bf results from local theory
  (\S\ref{local_approx})}. {\bf Note also that $k\simeq k_c$. }
Fig. \ref{2d_fargo_viz} shows the $m=1$
spiral is trailing, consistent with $k>0$.

@@

{\bf
  We can estimate the expected growth rate of the $m=1$ mode due to the
  temperature gradient. Setting $k = k_c$ and $m=1$ into
  Eq. \ref{theoretical_rate} gives
  \begin{align}\label{theoretical_rate1}
    \gamma \sim \frac{qh}{2Q}\Omega. 
  \end{align}
  Inserting $q=1$, $h=0.05$ and $Q\simeq 1.5$  gives
  $\gamma \simeq 0.017\Omega$, consistent with numerical
  results.  
}

\subsubsection{Angular momentum exchange with the background disc}  
{\bf We explicitly show that the growth of the $m=1$ spiral is due to
  the forced temperature gradient via the background torque described
  in \S\ref{wkb}. We integrate the statement for angular momentum 
  conservation for linear perturbations, Eq. \ref{lin_ang_mom_cons},
  assuming boundary fluxes are negligible, to obtain
}

\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density associated with the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We  
compute both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The average discrepancy is $\simeq 5\%$. 
The match is less good later on, when the spiral
amplitude is no longer small ($\mathrm{max}\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\mathrm{max}\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less applicable. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its {\bf (negative)} angular
momentum, {\bf leading to its amplitude growth.} This is consistent
with angular momentum component measurements (Fig. \ref {2d_angmom}).

>>>>

\subsection{Reference run}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc is  
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibly due to non-linear interaction of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

Fig. \ref{2d_angmom} shows the evolution of disc angular momentum
components. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum,  which indicates it is a
  low-frequency mode. 
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-armed spiral to the background disc.

@@

\subsection{Dependence on the imposed temperature profile}
 We show that the growth of the $m=1$ spiral is
  associated with the imposed temperature gradient  by performing a 
  series of simulations with $q\in[0,1]$.  However, to maintain similar Toomre $Q$
  profiles, we adjust the surface density power-law index such
  that $s = (3+q)/2$.  For clarity these simulations are initialised
  with $m=1$ perturbations only.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq}   
  \caption{Evolution of the $m=1$ spiral amplitude as a function of
    the negative of the imposed temperature gradient $q$. The maximum value of the
    $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
    \label{fargo_varq}} 
\end{figure}

We plot growth rates of the $m=1$ mode as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a 
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates 
 As the background torque is proportional to $q$
  (Eq. \ref{theoretical_rate}), 
this indicates that the temperature gradient is responsible for the
development of the one-armed spirals observed in our simulations.

@@

\subsection{Properties of the $m=1$ spiral and its growth}\label{fargo_m1}
 Here we analyse the $q=1$ case in Fig. \ref{fargo_varq_growth} in
  more detail.

Fig. \ref{2d_fargo_viz} shows a snapshot of the $m=1$ surface
density of this run. 
By measuring the $m=1$ surface density amplitude and its
pattern speed, we obtain a co-rotation radius and growth rate 
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
This one-armed spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We calculate $k$ numerically and plot
its normalised value in Fig. \ref{fargo_wavenumber}. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply results from local theory
  (\S\ref{local_approx}). Note also that $k\simeq k_c$. 
Fig. \ref{2d_fargo_viz} shows the $m=1$
spiral is trailing, consistent with $k>0$.

@@

  We can estimate the expected growth rate of the $m=1$ mode due to the
  temperature gradient. Setting $k = k_c$ and $m=1$ into
  Eq. \ref{theoretical_rate} gives
  \begin{align}\label{theoretical_rate1}
    \gamma \sim \frac{qh}{2Q}\Omega. 
  \end{align}
  Inserting $q=1$, $h=0.05$ and $Q\simeq 1.5$  gives
  $\gamma \simeq 0.017\Omega$, consistent with numerical
  results.

\subsubsection{Angular momentum exchange with the background disc}  
 We explicitly show that the growth of the $m=1$ spiral is due to
  the forced temperature gradient via the background torque described
  in \S\ref{wkb}. We integrate the statement for angular momentum 
  conservation for linear perturbations, Eq. \ref{lin_ang_mom_cons},
  assuming boundary fluxes are negligible, to obtain

\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density associated with the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We  
compute both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The average discrepancy is $\simeq 5\%$. 
The match is less good later on, when the spiral
amplitude is no longer small ($\mathrm{max}\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\mathrm{max}\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less applicable. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its (negative) angular
momentum, leading to its amplitude growth. This is consistent
with angular momentum component measurements (Fig. \ref {2d_angmom}).

### 2015-02-06 paper/summary.tex
first revision complete, but still need spell check and proof read
paper accepted, minor corrections

\section{Summary and conclusions}\label{summary}
{\bf
  In this paper, we have described a destabilising  
  effect of adopting a fixed temperature profile to model   
  astrophysical discs. By applying angular momentum conservation
  within linear theory, we showed that a forced temperature gradient 
  introduces a torque on linear perturbations. We call this the
  background torque because it represents an exchange of angular
  momentum between the background disc and the perturbations. This 
  offers a previously unexplored pathway to instability in locally
  isothermal discs.  
}

In the local approximation, we showed that this background torque is
negative for {\bf non-axisymmetric} trailing waves in discs with a fixed temperature or
sound-speed profile that decrease outwards. {\bf The background torque} enforces  
low-frequency {\bf non-axisymmetric} modes because they are associated
with negative angular momentum.

{\bf 
  We demonstrated the destabilising effect of the background torque by
  carrying out direct numerical hydrodynamic simulations of 
  locally isothermal discs with a self-gravitating surface density
  bump. 
}
We find such systems {\bf are} unstable to low-frequency perturbations
with azimuthal wavenumber $m=1$, which leads to the development of an one-armed 
trailing spiral that persist for at least $O(10^2)$ orbits. The spiral
pattern speed is smaller than the local disc rotation and
growth rates are $O(10^{-2}\Omega)$ which gives a characteristic
growth time of $O(10)$ orbits.

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to show that the growth of     
one-armed spirals in our disc model is due to the imposed
temperature {\bf gradient: growth rates increased linearly}  
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations. {\bf This 
  one-armed spiral instability can be interpreted as an initially 
  neutral, tightly-wound $m=1$ mode being destabilised by the 
  background torque.} The spiral
is mostly confined between two $Q$-barriers surface density bump. 
We find the instability behaves similarly in 2D and 3D {\bf, but in 3D}
the spiral disturbance becomes more radially global away from    
the midplane.

@@

{\bf There are several issues that remain to be addressed in future
  works:}

{\bf \emph{Thermal relaxation.} The locally isothermal assumption 
} can be relaxed by including an energy equation with
{\bf a source term that restores the disc temperature over a
  characteristic timescale $t_\mathrm{relax}$}. Preliminary FARGO
simulations indicate a thermal relaxation timescale $t_\mathrm{relax} <
0.1\Omega_k^{-1}$ is needed for the one-armed spiral to
develop. However, this value may {\bf be model-dependent} and will be
explored in a follow-up study. {\bf For example,
a longer $t_\mathrm{relax}$ may be permitted with larger temperature
gradients.}

@@

{\bf \emph{Non-linear evolution}.} In the deeply non-linear regime, the
one-armed spiral may  
shock and deposit negative angular momentum onto 
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
maintained, it may be possible to achieve a balance between the gain
of negative angular momentum through the background torque, and the
gain of positive angular momentum through shock dissipation. We remark  
that fragmentation {\bf of a low-frequency spiral} is unlikely
because the co-rotation radius is outside the bulk of the spiral arm, in the non-self-gravitating
portion of the disc \citep{durisen08}. In order
to study these possibilities, improved numerical models are needed to
ensure total angular momentum conservation on timescales much longer
than that considered in this paper.

@@

{\bf The background torque in other contexts.}  
{\bf In our simulations}, self-gravity coupled with the disc 
structure permits a neutral one-armed mode, which is then
destabilised by the {\bf fixed} temperature gradient.  
Other origins of the neutral one-armed spiral should be
investigated. One possibility is in Be star discs \citep{rivinius13},
for which one-armed oscillations may explain long-timescale variations
in their emission lines \citep[see e.g.][and references 
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
strictly isothermal disc models, so it would be interesting to explore
the effect of a temperature gradient on the stability of these
oscillations.

>>>>

\section{Summary and conclusions}\label{summary}

  In this paper, we have described a destabilising  
  effect of adopting a fixed temperature profile to model   
  astrophysical discs. By applying angular momentum conservation
  within linear theory, we showed that a forced temperature gradient 
  introduces a torque on linear perturbations. We call this the
  background torque because it represents an exchange of angular
  momentum between the background disc and the perturbations. This 
  offers a previously unexplored pathway to instability in locally
  isothermal discs.

In the local approximation, we showed that this background torque is
negative for non-axisymmetric trailing waves in discs with a fixed temperature or
sound-speed profile that decrease outwards. A negative background torque enforces  
low-frequency non-axisymmetric modes because they are associated
with negative angular momentum.

  We demonstrated the destabilising effect of the background torque by
  carrying out direct numerical hydrodynamic simulations of 
  locally isothermal discs with a self-gravitating surface density
  bump. 
We find such systems are unstable to low-frequency perturbations
with azimuthal wavenumber $m=1$, which leads to the development of an one-armed 
trailing spiral that persist for at least $O(10^2)$ orbits. The spiral
pattern speed is smaller than the local disc rotation and
growth rates are $O(10^{-2}\Omega)$ which gives a characteristic
growth time of $O(10)$ orbits.

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to show that the growth of     
one-armed spirals in our disc model is due to the imposed
temperature gradient: growth rates increased linearly   
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations. This 
  one-armed spiral instability can be interpreted as an initially 
  neutral, tightly-wound $m=1$ mode being destabilised by the 
  background torque. The spiral
is mostly confined between two $Q$-barriers in the surface density bump. 
We find the instability behaves similarly in 2D and 3D, but in 3D
the spiral disturbance becomes more radially global away from    
the midplane.

@@

 There are several issues that remain to be addressed in future
  works:

 \emph{Thermal relaxation.} The locally isothermal assumption 
 can be relaxed by including an energy equation with
 a source term that restores the disc temperature over a
  characteristic timescale $t_\mathrm{relax}$. Preliminary FARGO
simulations indicate a thermal relaxation timescale $t_\mathrm{relax} <
0.1\Omega_k^{-1}$ is needed for the one-armed spiral to
develop. However, this value is likely model-dependent. For example,
a longer $t_\mathrm{relax}$ may be permitted with larger temperature
gradients. This issue, together with a parameter survey, will be
considered in a follow-up study.

@@

 \emph{Non-linear evolution}. In the deeply non-linear regime, the
one-armed spiral may  
shock and deposit negative angular momentum onto 
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
maintained, it may be possible to achieve a balance between the gain
of negative angular momentum through the background torque, and the
gain of positive angular momentum through shock dissipation. We remark  
that fragmentation is unlikely because the co-rotation radius is
outside the bulk of the spiral arm \citep{durisen08,rogers12}. In order
to study these possibilities, improved numerical models are needed to
ensure total angular momentum conservation on timescales much longer
than that considered in this paper.

%interaction with turbulence

@@

\emph{Other applications of the background torque.}     
 The background torque is a generic feature in  
  discs for which the temperature is set externally.  It may therefore
  be relevant in other astrophysical contexts. 
One possibility is in Be star discs \citep{rivinius13}, 
for which one-armed oscillations may explain long-timescale variations
in their emission lines \citep[see e.g.][and references 
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies 
invoke alternative mechanisms to produce \emph{neutral} one-armed
oscillations (e.g. rotational deformation of the star), but consider 
strictly isothermal discs. It would be interesting to explore
the effect of a radial temperature gradient on the stability of these
oscillations.
